Shallow EDSLs and Object-Oriented Programming

### 2016-06-22 paper/sections/ShallowOO.tex

\paragraph{Porting to Scala}
Indeed, we can easily translate the Haskell program into an OO
language like Scala, as shown in Figure~\ref{code:base}.
The record type maps to the trait \lstinline{Circuit} and field
declaration becomes a method declaration.
Each case in the semantic function corresponds to an class and its parameters become fields of the class.
And all the classes extends \lstinline{Circuit} by implementing \lstinline{width}.

\begin{figure}
\lstinputlisting[linerange=2-15]{../src/width/Circuit.scala}%APPLY:CIRCUIT_SCALA
\caption{OO implementation of \dsl}
\label{code:base}
\end{figure}

>>>>

\paragraph{Porting to Scala}
Indeed, we can easily translate the Haskell program into an OO
language like Scala:
\lstinputlisting[linerange=2-15]{../src/width/Circuit.scala}%APPLY:CIRCUIT_SCALA
The record type maps to the trait \lstinline{Circuit} and field
declaration becomes a method declaration.
Each case in the semantic function corresponds to an class and its parameters become fields of the class.
And all the classes extends \lstinline{Circuit} by implementing \lstinline{width}.

### 2016-06-22 paper/sections/Transformations.tex

Figure~\ref{code:desugar} shows how to eliminate \lstinline{Id} construct in Scala.
\begin{figure}
\lstinputlisting[linerange=4-23]{../src/desugar/Circuit.scala}%APPLY:DESUGAR_SCALA
\caption{OO implementation of desugaring}
\label{code:desugar}
\end{figure}
The newly introduced method \lstinline{desugar} in \lstinline{Circuit} returns a
\lstinline{Circuit} reflecting the nature of a transformation. The implementation of \lstinline{desugar} for each case is
straightforward: \emph{fan n} returns itself; $beside\ c_1\ c_2$ recursively desugars $c_1$ and $c_2$ then wraps back
their results into an instance of \lstinline{Beside}; \emph{id n} just mimics the
formula.

%\weixin{define a show function in the and show the result after desugaring?}
Inspired by the Scala implementation, we can port it back to Haskell:
\lstinputlisting[linerange=6-21]{./code/desugar.hs}%APPLY:DESUGAR
which relies on the laziness of Haskell.

\subsection{Sophisticated Optimizations}
Sophisticated optimizations may need to inspect multiple internal
representations of values.
Suppose we want to merge consecutive vertical wires:
$$
beside\ (id\ m)\ (id\ n) = id\ (m + n)
$$
This rewrite rule applies only when both $c_1$ and $c_2$ of the $beside\ c_1\ c_2$
are $id$, which has to inspect the internal representation of $c_1$ and $c_2$ simultaneously.
However, pattern matching can only be simulated only at the the top level, i.e. $beside$.
There is no easy way to destruct $c_1$ and $c_2$.
Fortunately, Scala has built-in support for pattern matching. For extensibility reasons, we
can not simply decorate classes with \lstinline{case} modifier.
% as it is not possible to have a case class that extends another case class.
Alternatively, we manually define the \lstinline{unapply} method, a.k.a \emph{extractor}, in the companion object. Here comes the implementation
of the above optimization:
\lstinputlisting[linerange=4-22]{../src/optimizations/Circuit.scala}%APPLY:MERGEIDS_SCALA

For other OO languages, it is still possible to simulate
pattern matching through test methods or type test and type casts,
although they may not be as elegant as extractors~\cite{emir2007matching}.
Interested reader can refer to \ref{} for implementations using these approaches.

>>>>

The following shows how to eliminate \lstinline{Id} construct in Scala:
\lstinputlisting[linerange=4-23]{../src/desugar/Circuit.scala}%APPLY:DESUGAR_SCALA
The newly introduced method \lstinline{desugar} returns a
\lstinline{Circuit} reflecting the nature of transformations. The implementation of \lstinline{desugar} for each case is
straightforward: \emph{fan n} returns itself; $beside\ c_1\ c_2$ returns a new
$beside$ with $c_1$ and $c_2$ desugared; \emph{id n} just mimics the formula.

%\weixin{define a show function in the and show the result after desugaring?}
Inspired by the Scala implementation, we can port it back to Haskell thanks to
the laziness of Haskell:
\lstinputlisting[linerange=6-21]{./code/desugar.hs}%APPLY:DESUGAR

\subsection{Sophisticated Optimizations}
Sophisticated optimizations may need to inspect multiple internal
representations of values.
Suppose we want to merge consecutive vertical wires:
$$
beside\ (id\ m)\ (id\ n) = id\ (m + n)
$$
This rewrite rule applies only when both $c_1$ and $c_2$ of $beside\ c_1\ c_2$
are $id$ and also accesses their inner value.
This kind of optimizations are hard to define in shallow embeddings since it
requires deep pattern matching on the AST.
Fortunately, Scala has built-in support for pattern matching on objects.
We can not simply decorate classes with \lstinline{case} modifier
as case class to case class extension is not supported.
Alternatively, we could manually define an \lstinline{unapply} method, a.k.a
\emph{extractor}, in the companion object:
%Here comes the implementation of the optimization:
\lstinputlisting[linerange=4-22]{../src/optimizations/Circuit.scala}%APPLY:MERGEIDS_SCALA

For other OO languages, pattern matching can still be simulated through \emph{test methods} or \emph{type test and type casts},
although they may not be as elegant as extractors~\cite{emir2007matching}.
Interested reader can refer to \ref{} for implementations using these approaches.
